---
title: "Conway BUOW Band Table Cleaning
**2000 - 2019**"
author: "Matthew Dunning"
date: "2024-12-05"
output: html_document
---
# Cleaning Band Data 2000-2019
```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Cleaning Band

This section is to clean all burrowing data from 2000 - July 2024. This section is only focused on getting

```{r install r packages1, eval = FALSE, echo = TRUE}
install.packages("tidyverse")
install.packages("readxl")
install.packages()
```

## Load packages

load packages needed

```{r load libraries1, eval = FALSE, echo = TRUE}
library(tidyverse)
library(readxl)
```

First lets clean the data from 2000-2019. First lets import the data into r and then create a new data frame with only the needed columns.

```{r import 2000-2029 data, eval = FALSE, echo = TRUE}
setwd("~/BUOW_Migration-Maps/data")
BUOW1 <- read_csv("BUOW_Data2000-2019.csv")
bands <- select(BUOW1, band, USGSLeg, acraftColor, acraftFirstCode, acraftType, acraftSecondCode, AcraftLeg, right_top, right_bottom, left_top, left_bottom, sex)
```

## Add hyphens in the correct locations

First we need to add some criteria and conditions

```{r add several criteria and conditions, eval = FALSE, echo = TRUE}
#first need to add several criteria and conditions
criteria1 <- "0844"
criteria2 <- "0934" #some bands already had a dash so now there is --
criteria3 <- "804"
criteria4 <- "1204"
criteria5 <- "764"
criteria6 <- "544"
criteria7 <- "574"
criteria8 <- "594"
criteria9 <- "914"
criteria10 <- "1005"
criteria11 <- "934" #some bands already had a dash so now there is --
criteria12 <- "844"
Condition1 <- grepl("^AL-", bands$band)
criteria <- c(criteria1, criteria2, criteria4,  criteria10)
criteriax <- c(criteria11, criteria12, criteria3, criteria5, criteria6, criteria7,
               criteria8, criteria9)
```

Next run the code

```{r code to fix band hyphens, eval = FALSE, echo = TRUE}
practice_band <- bands %>%
  mutate(band = case_when(
    substr(band, 1, 4) %in% criteria ~ {
      # Remove existing hyphen if it exists
      clean_band <- gsub("-", "", band)
      # Ensure you're using the clean version
      paste0(substr(clean_band, 1, 4), "-", substr(clean_band, 5, nchar(clean_band)))
    },
    TRUE ~ band  # Keep original band if no criteria match
  )) %>% 
  mutate(band = case_when(
    substr(band, 1, 3) %in% criteriax ~ {
      # Remove existing hyphen if it exists
      clean_band <- gsub("-", "", band)
      # Ensure you're using the clean version
      paste0(substr(clean_band, 1, 3), "-", substr(clean_band, 4, nchar(clean_band)))
    },
    TRUE ~ band # Keep original band if no criteria match
  )) %>%
  mutate(band =ifelse(Condition1, substr(band, 4, nchar(band)), band)
         )
```

remove the criteria and conditions to keep the r environment tidy

```{r remove criteria and conditions for band hyphens, eval = FALSE, echo = TRUE}
remove(criteria, criteria1, criteria2, criteria3, criteria4, criteria5, criteria6,
       criteria7, criteria8, criteria9, criteria10, criteria11, criteria12,
       criteriax, Condition1)
```

## Fix some capitalization issues

```{r fix spelling in acraft and left_top, eval = FALSE, echo = TRUE}
practice_band <- practice_band %>%
  mutate(acraftSecondCode = str_replace_all(acraftSecondCode, "w", "W")
         )

practice_band <- practice_band %>%
  mutate(left_top = str_replace_all(left_top, "E", "e")
         )
```

## Proofing resights
This code can be used to proof the resights

```{r proofing re-sights, eval = FALSE, echo = TRUE}
columns_to_check <- c("USGSLeg", "acraftColor", "acraftFirstCode", 
                      "acraftType", "acraftSecondCode", "AcraftLeg", 
                      "right_top", "right_bottom", "left_top", 
                      "left_bottom"
                      )
repeating_values <- practice_band %>%
  group_by(band) %>%
  filter(any(sapply(columns_to_check, function(col) n_distinct(get(col)) > 1))) %>% 
  ungroup()
#values that have contradictory info about band position should be edited
#values that are U and M/F/J/UA should default to M/F
```

## Clean the sex column

```{r Change the Sex column,  eval = FALSE, echo = TRUE}
#First make all other values U
practice_band <- practice_band %>%
  mutate(
    sex = case_when(
      sex %in% c("J", "UA") ~ "U",  # Replace "J" and "UA" with "U"
      is.na(sex) ~ "U",             # Replace NA with "U"
      TRUE ~ sex                    # Keep all other values unchanged
    )
  )

#Make all bands that contain both M and F into C. This also makes U a C in these cases.
bands_with_both <- practice_band %>%
  filter(sex %in% c("M", "F")) %>%  # Filter rows with "M" or "F"
  group_by(band) %>%                # Group by band
  filter(n_distinct(sex) == 2) %>%  # Ensure the band contains both "M" and "F"
  ungroup() %>%                     # Ungroup after filtering
  pull(band)                        # Extract the band names

# Step 2: Replace "M", "F", and "U" with "C" only for bands that contain both "M" and "F"
practice_band <- practice_band %>%
  mutate(
    sex = case_when(
      band %in% bands_with_both & sex %in% c("M", "F", "U") ~ "C",  # Replace "M", "F", "U" with "C" if band has both "M" and "F"
      TRUE ~ sex  # Leave all other values unchanged
    )
  )

#changes MU and FU combos to M or F
bands_with_m_u <- practice_band %>%
  filter(sex %in% c("M", "U")) %>%  # Focus on "M" and "U"
  group_by(band) %>%
  filter(any(sex == "M") & any(sex == "U")) %>%  # Band must have both "M" and "U"
  ungroup() %>%
  pull(band)  # Extract the band names

bands_with_f_u <- practice_band %>%
  filter(sex %in% c("F", "U")) %>%  # Focus on "F" and "U"
  group_by(band) %>%
  filter(any(sex == "F") & any(sex == "U")) %>%  # Band must have both "F" and "U"
  ungroup() %>%
  pull(band)  # Extract the band names

# Step 2: Mutate the sex column to replace "U" with "M" or "F" based on the logic above
practice_band <- practice_band %>%
  mutate(
    sex = case_when(
      band %in% bands_with_m_u & sex == "U" ~ "M",  # If band has both "M" and "U", change "U" to "M"
      band %in% bands_with_f_u & sex == "U" ~ "F",  # If band has both "F" and "U", change "U" to "F"
      TRUE ~ sex  # Otherwise, keep the original value of sex
    )
  )
```

## Now clean the USGS band info

```{r Now changing the USGSLeg, eval = FALSE, echo = TRUE}
# Update the USGSLeg values to "C" if both "Right" and "Left" are present in the same band
bands_to_change <- practice_band %>%
  group_by(band) %>%
  filter(any(USGSLeg == "Right") & any(USGSLeg == "Left")) %>%
  pull(band)  # Extract the band names that meet the condition

# Now mutate 'USGSLeg' to "C" for the identified bands
practice_band <- practice_band %>%
  mutate(
    USGSLeg = if_else(
      band %in% bands_to_change,  # Check if the band is in the list of bands to change
      "C",  # Change to "C"
      USGSLeg  # Otherwise, keep original value
    )
  )

#change the NA to right/left in USGSLeg
bands_to_fix <- practice_band %>%
  group_by(band) %>%
  filter(any(is.na(USGSLeg)) & any(USGSLeg == "Right")) %>%  # Check for bands with both NA and "Right"
  ungroup() %>%
  pull(band)  # Extract the band names that meet the condition

# Step 2: Apply the changes to 'USGSLeg' for these identified bands
practice_band <- practice_band %>%
  group_by(band) %>%
  mutate(
    USGSLeg = case_when(
      band %in% bands_to_fix & is.na(USGSLeg) & any(USGSLeg == "Right") ~ "Right",  # Replace NA with "Right" if "Right" exists in the band
      TRUE ~ USGSLeg  # Keep other values unchanged
    )
  ) %>%
  ungroup()  # Ungroup after applying the update

#Now the Left Leg
bands_to_fix <- practice_band %>%
  group_by(band) %>%
  filter(any(is.na(USGSLeg)) & any(USGSLeg == "Left")) %>%  # Check for bands with both NA and "Left"
  ungroup() %>%
  pull(band)  # Extract the band names that meet the condition

# Step 2: Apply the changes to 'USGSLeg' for these identified bands
practice_band <- practice_band %>%
  group_by(band) %>%
  mutate(
    USGSLeg = case_when(
      band %in% bands_to_fix & is.na(USGSLeg) & any(USGSLeg == "Left") ~ "Left",  # Replace NA with "Left" if "Left" exists in the band
      TRUE ~ USGSLeg  # Keep other values unchanged
    )
  ) %>%
  ungroup()  # Ungroup after applying the update
```

## ACRAFT specific cleaning

First we only want to pull the bands that have ACRAFT bands associated with them.

```{r pull acraft bands from bands, eval = FALSE, echo = TRUE}
#Lets make all NA values say "no auxiliary band"
bands_to_fix <- practice_band %>%
  group_by(band) %>%
  filter(any(acraftType == "no auxiliary band") & any(is.na(acraftType))) %>%
  pull(band)

# Apply the change to the acraftType column
practice_band <- practice_band %>%
  group_by(band) %>%
  mutate(
    acraftType = case_when(
      band %in% bands_to_fix & is.na(acraftType) ~ "no auxiliary band",  # Replace NA with "no auxiliary band" where necessary
      TRUE ~ acraftType  # Keep other values unchanged
    )
  ) %>%
  ungroup()  # Remove grouping after applying the changes
```

```{some facts about edits, echo = FALSE}
#there are only 3 bands found that started as unbanded and were given a acraft band late.
#there are several acraft birds that do not have USGS bands.
#there are several birds from umatilla that were also no band, 2017.
#939-90924 is should be 934-90924 fixed within excel sheet. MD 10/26/24.
#0914-58717 had to be removed and replaced as 934-78740. MD 10/26/24
#0914-56412 had to be removed and replaced as 0914-56414. also had to change the year from 2015 to 2014
# actually the previous issue could not be resolved so is set aside both bands. MD 10/26/24
```

```{r acraft export, eval = FALSE, echo = TRUE}
#filter only the columns we want for ACRAFT bands
acraft <- practice_band %>% 
  select(band, USGSLeg, acraftColor, acraftFirstCode, acraftType, acraftSecondCode, AcraftLeg, sex) %>% 
  filter(acraftType != "no auxilary") %>% 
  filter(acraftType != "no auxiliary band") %>% 
  filter(acraftType != "plastic") %>% 
  filter(!is.na(acraftType))

#Remove all duplicate bands from recapture events
distinct_bands_all_columns <- acraft %>%
  distinct(band, USGSLeg, acraftColor, acraftFirstCode, acraftType,
           acraftSecondCode, AcraftLeg, sex)

acraft <- distinct_bands_all_columns %>%
  mutate(left_leg = case_when(
    grepl("Left", AcraftLeg, ignore.case = TRUE) ~ "Acraft",    # If "Left" in AcraftLeg, set 'Acraft'
    grepl("Left", USGSLeg, ignore.case = TRUE) ~ "USGS",         # If "Left" in USGSLeg, set 'USGS'
    is.na(USGSLeg) ~ NA_character_,                              # Leave NA as NA if USGSLeg is NA
    TRUE ~ as.character(USGSLeg)                                  # Leave other values unchanged (like "Right")
  ))

acraft <- acraft %>%
  mutate(right_leg = case_when(
    grepl("Right", AcraftLeg, ignore.case = TRUE) ~ "Acraft",    # If "Right" in AcraftLeg, set 'Acraft'
    grepl("Right", USGSLeg, ignore.case = TRUE) ~ "USGS",         # If "Right" in USGSLeg, set 'USGS'
    is.na(USGSLeg) ~ NA_character_,                              # Leave NA as NA if USGSLeg is NA
    TRUE ~ as.character(USGSLeg)                                  # Leave other values unchanged (like "Left")
  ))

#export the acraft bands as a data frame to be added to database
export_acraft <- acraft %>% 
  rename(
    BUOW_USGS_ID = band,
    ACRAFT_color = acraftColor,
    ACRAFT_first_code = acraftFirstCode,
    ACRAFT_second_code = acraftSecondCode,
    ACRAFT_layout = acraftType
  )

export_acraft <- export_acraft %>% 
  select(
    BUOW_USGS_ID,                  
    left_leg,               
    right_leg,            
    ACRAFT_color,           
    ACRAFT_first_code,       
    ACRAFT_layout,            
    ACRAFT_second_code,   
    sex,
  )
```

## Color band sorting

```{r color band sorting code, eval = FALSE, echo = TRUE}
#first select the columns needed for color bands
color_bands <- practice_band %>% 
  select(band, USGSLeg, acraftType, right_top, right_bottom, left_top, left_bottom, sex) %>% 
  filter(acraftType != "no auxilary") %>% 
  filter(acraftType != "no auxiliary band") %>% 
  filter(acraftType != "plastic-numeric") %>% 
  filter(acraftType != "bar") %>% 
  filter(acraftType != "normal") %>% 
  filter(acraftType != "over") %>% 
  filter(acraftType != "vbar") %>% 
  filter(!is.na(acraftType)
         )

#next we need to exclude recaptures and get distinct band numbers
color_bands <- color_bands %>%
  distinct(band, USGSLeg, acraftType, right_top, right_bottom, left_top,
           left_bottom, sex)

#next renambe the columns to match the database in SQL
export_color <- color_bands %>% 
  rename(
    BUOW_USGS_ID = band,
    top_left_band = left_top,
    bottom_left_band = left_bottom,
    top_right_band = right_top,
    bottom_right_band = right_bottom
  )

#finally pick all columns needed for export
export_color <- export_color %>% 
  select(BUOW_USGS_ID,
         top_left_band,
         bottom_left_band,
         top_right_band,
         bottom_right_band,
         sex)

```

## Get the rest of the bands that have neither ACRAFT nor Color bands

```{r export all other bands, eval = FALSE, echo = TRUE}
#select all columns needed for all other bands
all_other_bands <- practice_band %>%
  select(band, USGSLeg, acraftColor, acraftFirstCode, acraftType, acraftSecondCode, AcraftLeg, sex) %>%
  filter(!acraftType %in% c("plastic", "plastic-numeric", "bar", "normal", "over", "vbar") | is.na(acraftType))

#remove all recaptures and only get distinct band numbers
all_other_bands <- all_other_bands %>%
  distinct(band, USGSLeg, acraftColor, acraftFirstCode, acraftType, acraftSecondCode, AcraftLeg, sex)

#code that assigns left or right to indicate which leg the USGS band is placed
all_other_bands <- all_other_bands %>%
  mutate(
    left_leg = case_when(
      grepl("Left", USGSLeg, ignore.case = TRUE) ~ "USGS",         # If "Left" in USGSLeg, set 'USGS'
      grepl("Left", AcraftLeg, ignore.case = TRUE) ~ "USGS",        # If "Left" in AcraftLeg, set 'USGS'
      TRUE ~ NA_character_                                         # Otherwise, set 'NA'
    ),
    right_leg = case_when(
      grepl("Right", USGSLeg, ignore.case = TRUE) ~ "USGS",        # If "Right" in USGSLeg, set 'USGS'
      grepl("Right", AcraftLeg, ignore.case = TRUE) ~ "USGS",       # If "Right" in AcraftLeg, set 'USGS'
      TRUE ~ NA_character_                                         # Otherwise, set 'NA'
    )
  )

#rename columns to match SQL database
export_all_other_bands <- all_other_bands %>% 
  rename(BUOW_USGS_ID = band)
    
#select all columns for export to SQL    
export_all_other_bands <- export_all_other_bands %>% 
  select(
    BUOW_USGS_ID,                  
    left_leg,               
    right_leg,
    sex
  )
```

## Finally remove all unneccessary dataframes and objects to keep environment tidy

```{r remove objects and values, eval = FALSE, echo = TRUE}
remove(distinct_bands_all_columns, all_other_bands, test, acraft, bands, BUOW1, color_bands, repeating_values)
remove(bands_to_change, bands_to_fix, bands_with_both, bands_with_f_u, bands_with_m_u, columns_to_check)
remove(BUOW1, bands, color_bands, distinct_bands_all_columns, practice_band, repeating_values, test, acraft, all_other_bands)
```
