---
title: "Conway BUOW Band Table Cleaning
**2020 - August 2024**"
author: "Matthew Dunning"
date: "2024-12-06"
output: html_document
---
# Cleaning Band Data 2020 - August 2024
```{r setup2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install and load packages

```{r install packages2, eval=FALSE, echo=TRUE}
install.packages("tidyverse")
install.packages("readxl")
```

```{r load packages2, eval=TRUE, echo=TRUE}
library(tidyverse)
library(readxl)
```

## Set directory and load data

```{r set directory2, eval=TRUE, echo=TRUE}
 setwd("~/BUOW_Migration-Maps/data")
BUOW <- read_excel("RecentBUOWDataAdditions.xlsx")
```

Now select columns needed for cleaning

```{r select columns for cleaning, eval=TRUE, echo=TRUE}
bands <- select(BUOW, band, right_top, right_bottom, left_top, left_bottom,
                BandCombo, sex)
```

## Add hyphens where needed

```{r add criteria, eval=TRUE, echo=TRUE}
criteria1 <- "934" #some bands already had a dash so now there is --
criteria2 <- "914"
criteria3 <- "804"
criteriax <- c(criteria1, criteria2, criteria3)

practice_band <- bands%>% 
mutate(band = case_when(
  substr(band, 1, 3) %in% criteriax ~ {
    # Remove existing hyphen if it exists
    clean_band <- gsub("-", "", band)
    # Ensure you're using the clean version
    paste0(substr(clean_band, 1, 3), "-", substr(clean_band, 4, nchar(clean_band)))
  },
  TRUE ~ band)) # Keep original band if no criteria match

remove(criteria1, criteria2, criteria3, criteriax)
```

Before moving on we want to add a unique key to each capture event

```{r add a unique key, eval=TRUE, echo=TRUE}
practice_band <- practice_band %>% 
  mutate(unique_key = row_number()) %>%     # Adds a sequence number starting from 1
  select(unique_key, everything()) 
```

## Move colors into correct columns
First extract the colors and place in correct columns
```{r extract color combos, eval=TRUE, echo=TRUE}
practice_band <- practice_band %>%
  mutate(
    top_left_band = str_extract(BandCombo, "^[^/]+"),                          # Extract top left part (before first "/")
    bottom_left_band = str_extract(BandCombo, "(?<=/)[^:]+"),                   # Extract bottom left part (between "/" and ":")
    top_right_band = str_extract(BandCombo, "(?<=:)[^/]+"),                     # Extract top right part (between ":" and "/")
    bottom_right_band = str_extract(BandCombo, "(?<=/)[^/]+$")                  # Extract bottom right part (after last "/")
  )
```
Next only select bands that have a color combo
```{r select color combos, echo=TRUE, eval=TRUE}
practice_band <- practice_band %>% 
  mutate(right_top = ifelse(is.na(right_top), top_right_band, right_top),
         right_bottom = ifelse(is.na(right_bottom), bottom_right_band, right_bottom),
         left_top = ifelse(is.na(left_top), top_left_band, left_top),
         left_bottom = ifelse(is.na(left_bottom), bottom_left_band, left_bottom))
```
Next change the abbreviations to full spelling of colors
```{r spell out abbreviations, eval=TRUE, echo=TRUE}
practice_band <- practice_band %>%
  mutate(
    left_top = gsub("Re", "Red", left_top),
    left_top = gsub("NB", "No Band", left_top),
    left_top = gsub("Bk", "Black", left_top),
    left_top = gsub("Wh", "White", left_top),
    left_top = gsub("Ye", "Yellow", left_top),
    left_top = gsub("MBl", "Medium Blue", left_top),
    left_top = gsub("MGr", "Medium Green", left_top),
    left_top = gsub("Or", "Orange", left_top),
    left_top = gsub("LBl", "Light Blue", left_top),
    left_top = gsub("Gy", "Gray", left_top),
    left_top = gsub("AL", "USGS", left_top)
  )

practice_band <- practice_band %>%
  mutate(
    right_top = gsub("Re", "Red", right_top),
    right_top = gsub("NB", "No Band", right_top),
    right_top = gsub("Bk", "Black", right_top),
    right_top = gsub("Wh", "White", right_top),
    right_top = gsub("Ye", "Yellow", right_top),
    right_top = gsub("MBl", "Medium Blue", right_top),
    right_top = gsub("MGr", "Medium Green", right_top),
    right_top = gsub("Or", "Orange", right_top),
    right_top = gsub("LBl", "Light Blue", right_top),
    right_top = gsub("Gy", "Gray", right_top),
    right_top = gsub("AL", "USGS", right_top)
  )

practice_band <- practice_band %>%
  mutate(
    left_bottom = gsub("Re", "Red", left_bottom),
    left_bottom = gsub("NB", "No Band", left_bottom),
    left_bottom = gsub("Bk", "Black", left_bottom),
    left_bottom = gsub("Wh", "White", left_bottom),
    left_bottom = gsub("Ye", "Yellow", left_bottom),
    left_bottom = gsub("MBl", "Medium Blue", left_bottom),
    left_bottom = gsub("MGr", "Medium Green", left_bottom),
    left_bottom = gsub("Or", "Orange", left_bottom),
    left_bottom = gsub("LBl", "Light Blue", left_bottom),
    left_bottom = gsub("Gy", "Gray", left_bottom),
    left_bottom = gsub("ACGr", "ACRAFT Green", left_bottom),
    left_bottom = gsub("AL", "USGS", left_bottom)
  )

practice_band <- practice_band %>%
  mutate(
    right_bottom = gsub("Re", "Red", right_bottom),
    right_bottom = gsub("NB", "No Band", right_bottom),
    right_bottom = gsub("Bk", "Black", right_bottom),
    right_bottom = gsub("Wh", "White", right_bottom),
    right_bottom = gsub("Ye", "Yellow", right_bottom),
    right_bottom = gsub("MBl", "Medium Blue", right_bottom),
    right_bottom = gsub("MGr", "Medium Green", right_bottom),
    right_bottom = gsub("Or", "Orange", right_bottom),
    right_bottom = gsub("LBl", "Light Blue", right_bottom),
    right_bottom = gsub("Gy", "Gray", right_bottom),
    right_bottom = gsub("AL", "USGS", right_bottom)
  )
```
Remove some columns that are no longer needed
```{r remove band columns not needed, eval=FALSE, echo=TRUE}
practice_band <- select(practice_band,
                        -BandCombo,
                        -top_right_band,
                        -top_left_band,
                        -bottom_right_band,
                        -bottom_left_band)
```

## Proof and Edit the sex column
First standardize all variables in the sex column
```{r standardize sex columns, eval=TRUE, echo=TRUE}
practice_band <- practice_band %>% 
  mutate(sex = gsub("female", "F", sex),
         sex = gsub("male", "M", sex),
         sex = gsub("unknown", "U", sex),
         sex = gsub("Unk", "U", sex),
         sex = gsub("J", "U", sex),
         sex = gsub("hy", "U", sex),
         sex = gsub("X", NA, sex))
```

There are a few cases where there are conflicting sex variables for the same bird. There can only be one value in the band table so all
these values need to be changed to C to reflect this conflict.

```{r filter and change all conflicting sex variables, eval=TRUE, echo=TRUE}
#Make all bands that contain both M and F into C. This also makes U a C in these cases.
bands_with_both <- practice_band %>%
  filter(sex %in% c("M", "F")) %>%  # Filter rows with "M" or "F"
  group_by(band) %>%                # Group by band
  filter(n_distinct(sex) == 2) %>%  # Ensure the band contains both "M" and "F"
  ungroup() %>%                     # Ungroup after filtering
  pull(band)                        # Extract the band names

# Step 2: Replace "M", "F", and "U" with "C" only for bands that contain both "M" and "F"
practice_band <- practice_band %>%
  mutate(
    sex = case_when(
      band %in% bands_with_both & sex %in% c("M", "F", "U") ~ "C",  # Replace "M", "F", "U" with "C" if band has both "M" and "F"
      TRUE ~ sex  # Leave all other values unchanged
    )
  )

```

There are some cases where a bird was caught as a HY so sex was unknown and it was later
caught as an adult and the sex was known. All these U values should be changed to M or F
because there can only be one value per band.

```{r change unknown to male or female, eval=TRUE, echo=TRUE}
#changes MU and FU combos to M or F
bands_with_m_u <- practice_band %>%
  filter(sex %in% c("M", "U")) %>%  # Focus on "M" and "U"
  group_by(band) %>%
  filter(any(sex == "M") & any(sex == "U")) %>%  # Band must have both "M" and "U"
  ungroup() %>%
  pull(band)  # Extract the band names

bands_with_f_u <- practice_band %>%
  filter(sex %in% c("F", "U")) %>%  # Focus on "F" and "U"
  group_by(band) %>%
  filter(any(sex == "F") & any(sex == "U")) %>%  # Band must have both "F" and "U"
  ungroup() %>%
  pull(band)  # Extract the band names

# Step 2: Mutate the sex column to replace "U" with "M" or "F" based on the logic above
practice_band <- practice_band %>%
  mutate(
    sex = case_when(
      band %in% bands_with_m_u & sex == "U" ~ "M",  # If band has both "M" and "U", change "U" to "M"
      band %in% bands_with_f_u & sex == "U" ~ "F",  # If band has both "F" and "U", change "U" to "F"
      TRUE ~ sex  # Otherwise, keep the original value of sex
    )
  )

```
## Export data and remove objects
```{r export recent BUOW Band data, eval=TRUE, echo=TRUE}
export_recentBUOWtable <- select(practice_band,
                 BUOW_USGS_ID = band,
                 top_left_band = left_top,
                 bottom_left_band = left_bottom,
                 top_right_band = right_top,
                 bottom_right_band = right_bottom,
                 sex)

remove(bands, BUOW, practice_band, bands_with_both, bands_with_f_u, bands_with_m_u)
```

