---
title: "Database Cleaning Captures Table
**Doubled Entried Rows**"
author: "Matthew Dunning"
date: "2024-12-06"
output: html_document
---
## Cleaning dobule entried rows
```{r setup4, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Import Data
This is a continuation of the from the previous chapter. This code is to seperate the multiple
capture evetns in a row to make new rows.
```{r load data4, eval=TRUE, echo=TRUE}
BUOW1 <- read_csv("C:/Users/mldunning/Documents/BUOW_Migration-Maps/data/BUOW_Data2000-2019.csv")
```
We need to select the rows with that have multiple entries. We then make two seperate
data frame to prepare the data.
```{r select data4, eval=TRUE, echo=TRUE}
buow1_with_date <- BUOW1 %>%
  filter(!is.na(meas_date_1))  # Keep only rows where 'meas_date_1' is not NA
  
row1 <- buow1_with_date
row2 <- buow1_with_date

row1 <- select(row1, 
               -`meas_date_2`, 
               -`est_age_2_(days)`, 
               -`tarsometatarsus_length_2_(mm)`,
               -`weight_2_(g)`,
               -`wing_2_(mm)`,
               -`tail_2_(mm)`,
               -`parasites_2`,
               -`comments...103`,
               -eggs,
               -chicks)
row1 <- row1 %>% 
  mutate(unique_key = row_number())
row1 <- row1 %>% 
  rename(
    meas_date = `meas_date_1`,
    est_age = `est_age_1_(days)`,
    tarsometatarsus_length = `tarsometatarsus_length_1_(mm)`,
    weight_g = `weight_1_(g)`,
    wing_mm = `wing_1_(mm)`,
    tail_mm = `tail_1_(mm)`,
    parasites = `parasites_1`,
    comments = `comments...93`
  )

row2 <- select(row2,
               -`meas_date_1`,
               -`est_age_1_(days)`,
               -`tarsometatarsus_length_1_(mm)`,
               -`weight_1_(g)`,
               -`wing_1_(mm)`,
               -`tail_1_(mm)`,
               -`parasites_1`,
               -`comments...93`,
               -eggs,
               -chicks)
row2 <- row2 %>%
  mutate(unique_key = row_number() + 4999)
row2 <- row2 %>% 
  rename(
    meas_date = `meas_date_2`,
    est_age = `est_age_2_(days)`,
    tarsometatarsus_length = `tarsometatarsus_length_2_(mm)`,
    weight_g = `weight_2_(g)`,
    wing_mm = `wing_2_(mm)`,
    tail_mm = `tail_2_(mm)`,
    parasites = `parasites_2`,
    comments = `comments...103`
  )
```
This code is to turn specific columns to numeric for future cleaning and filtering
```{r turn columns numeric, eval=TRUE, echo=TRUE}
row1 <- row1 %>%
  mutate_at(vars(tarsometatarsus_length, weight_g, wing_mm, tail_mm, parasites), as.numeric)

row2 <- row2 %>%
  mutate_at(vars(tarsometatarsus_length, weight_g, wing_mm, tail_mm, parasites), as.numeric)

row2$est_age <- as.character(row2$est_age)

#Now combine the rows into a single data frame.
combined_rows <- bind_rows(row1, row2)
```
### Clean data
```{r code for cleaning, eval=TRUE, echo=TRUE}
#clean HY days
juvage <- select(combined_rows,
                 unique_key,
                 est_age)

juvage <- juvage %>% 
  mutate(est_age = gsub(" days", "", est_age),
         est_age = gsub("not found", NA, est_age),
         est_age = gsub("not present", NA, est_age),
         est_age = gsub("UNB", NA, est_age),
         est_age = gsub("(\\d+)\\+", ">\\1", est_age),
         est_age = gsub(" ", "", est_age),
         est_age = gsub("6orless", "<6", est_age)
  )
juvage <- select(juvage,
                 unique_key,
                 "juvenile_estimated_age(days)" = est_age)

combined_rows <- combined_rows %>% 
  left_join(juvage, by = "unique_key")

combined_rows <- select(combined_rows,
                   -est_age)


#clean site names
combined_rows <- combined_rows %>% 
  mutate(
    site = gsub("TEAD", "Tooele Army Depot", site),
    site = gsub("UAD", "Umatilla Army Depot", site),
    site = gsub("SSNWR", "Salton Sea National Wildlife Refuge", site)
  )
```
## Prepare data for export
```{r export data, eval=TRUE, echo=TRUE}
exportumatillacaptures <- select(combined_rows,
                 band,
                 date = `meas_date`,
                 site,
                 country = Country,
                 state = State,
                 "juvenile_estimated_age(days)",
                 number_of_ectoparasites_categorical = parasites,
                 weight_g,
                 wing_cord_mm = wing_mm,
                 tail_length_mm = tail_mm,
                 tarsus_mm = tarsometatarsus_length,
                 comments)

exportumatillacaptures$date <- as.Date(exportumatillacaptures$date, format = "%d-%b-%y")
exportumatillacaptures$number_of_ectoparasites_categorical <- as.character(exportumatillacaptures$number_of_ectoparasites_categorical)


remove(BUOW1, buow1_with_date, combined_rows, juvage, row1, row2)

```

