---
title: "Combining export files to send to SQL"
author: "Matthew Dunning"
date: "2024-12-07"
output: html_document
---

# Import BUOW data into SQLlite

```{r setup7, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Combine All Band Tables

```{r combine all export band, eval=TRUE, echo=TRUE}
combineallbands <- bind_rows(export_acraft, export_all_other_bands, export_color,
                         export_recentBUOWtable)
```

We only want to select unique bands to import to sql database

```{r only select unique bands, eval=TRUE, echo=TRUE}
unique_combineallbands <- combineallbands[!duplicated(combineallbands$BUOW_USGS_ID), ]
```

```{R fix issue for import, eval=TRUE, echo=TRUE}
unique_combineallbands <- unique_combineallbands %>%
  mutate(left_leg = ifelse(tolower(left_leg) == "acraft", "ACRAFT", left_leg))

unique_combineallbands <- unique_combineallbands %>%
  mutate(right_leg = ifelse(tolower(right_leg) == "acraft", "ACRAFT", right_leg))

unique_combineallbands <- unique_combineallbands %>%
  mutate(left_leg = ifelse(left_leg == "Right", NA, left_leg))

unique_combineallbands <- unique_combineallbands %>%
  mutate(right_leg = ifelse(right_leg == "Left", NA, right_leg))

unique_combineallbands <- unique_combineallbands[!is.na(unique_combineallbands$BUOW_USGS_ID), ]
```

## Combine All Capture Tables

```{r combine all export captures, eval=TRUE, echo=TRUE}
combineallcaptures <- bind_rows(exportcaptures2020_2024, exportcapturesall, exportumatillacaptures)
```

Just need to make a quick units conversion before importing into the database

```{r convert elevation ft to meters, eval=TRUE, echo=TRUE}
combineallcaptures <- combineallcaptures %>%
  mutate(elevation_m = elevation_ft * 0.3048)

combineallcaptures <- combineallcaptures %>% 
  select(-elevation_ft)
```

We need to rename the columns for the import step to work later

```{r rename capture columns, eval=TRUE, echo=TRUE}
combineallcaptures <- combineallcaptures %>% 
  rename(
    BUOW_USGS_ID = band,
    right_wing_cord_mm = righ_wing_cord_mm,
    left_metatarsus_mm = left_tarsus_mm,
    right_metatarsus_mm = right_tarsus_mm,
    culmen_mm = culmen,
    P9_length_mm = P9Length_mm
  )


```

```{r delete NA, eval=TRUE, echo=TRUE}
combineallcaptures <- combineallcaptures[!is.na(combineallcaptures$BUOW_USGS_ID), ]
```

## Load Sites table

```{r load sites, eval=FALSE, echo=TRUE}
sites <- read_excel("C:/Users/mldunning/Documents/BUOW_Migration-Maps/data/sites_kmo.xlsx")

```

## Start import to SQL

```{r establish connection, eval=FALSE, echo=TRUE}
Burrowing_Owls <- dbConnect(RSQLite::SQLite(), "Burrowing_Owls.db")
```

### Import band table

First import the Band table. Every value has to match the set constratints and there can not be any repeating band numbers.

```{r write band table, eval=FALSE, echo=TRUE}
dbWriteTable(Burrowing_Owls, "Burrowing_Owls", unique_combineallbands, append = TRUE, row.names = FALSE)
```

### Import Capture Table

Next import the captures data. To do this we have to do a few extra steps because there are several unknown/unbanded birds that do not have a corresponding value in the band table. A foreign key is required to mainain data intergertiy.

```{r write the captures table, eval=FALSE, echo=TRUE}
dbWriteTable(Burrowing_Owls, "captures", combineallcaptures, append = TRUE, row.names = FALSE)
```

### Import Site Table
```{r write the sites table, eval=FALSE, echo=TRUE}
dbWriteTable(Burrowing_Owls, "sites", sites, append = TRUE, row.names = FALSE)
```

